---
title: "R Assignment"
subtitle: "by Sarah Phelps"
date: "March 7, 2024"
author: "Sarah Phelps"
root.dir: "/Users/sarahphelps/desktop/BCB546/BCB546/"
format:
  html:
    theme: cosmo 
    toc: true
    number-depth: 3
    html-math-method: katex
    code-fold: false
    page-layout: full
    embed-resources: true     
    standalone: true
editor: source
---

```{r}
#| label: knitr_setup
#| warning: false
#| echo: true

knitr::opts_chunk$set(echo       = TRUE,
                      warning    = TRUE,
                      error      = TRUE,
                      fig.height = 6,
                      fig.width  = 8
                      )


```


```{r}
fang_file <- paste0(getwd(), "/fang_et_al_genotypes.txt")
snp_file <- paste0(getwd(), "/snp_position.txt")

# load libraries
library(tidyverse)

# Read in data files
fang_data <- read.table(fang_file, header = T, sep = "\t")

snp_data <- read.table(snp_file, header = T, sep = "\t")

```


# Data Inspection
```{r}
# Fang data

#fang_head <- head(fang_data)
# We can see there are 986 columns in the Fang genotypes data set
fang_rows <- nrow(fang_data)
# there are 2782 rows in this data set
fang_size <- file.info("fang_et_al_genotypes.txt")
# file size = 11051939 bytes

# snp_head <- head(snp_data)
# We can see there are 15 colums in the SNP data set
snp_rows <- nrow(snp_data)
# there are 983 rows in this data set
snp_size <- file.info("snp_position.txt")
#file size = 82763 bytes

fang_head
fang_rows
fang_size 

#snp_head its takes up too much space but i ran it to see column names and format
snp_rows
snp_size 
```



# Data Processing of Fang Data
```{r}
# Processing Maize

# filter ZMMIL, ZMMLR, and ZMMMR
maize_subset <- subset(fang_data, Group %in% c("ZMMIL", "ZMMLR", "ZMMMR"))

# get rid of column 1 and 2
clean_maize <- select(maize_subset, abph1.20:zmm3.4)

trans_maize <- data.frame(t(clean_maize))


# Processing Teosinte

# filter ZMPBA, ZMPIL, ZMPJA
teos_subset<- subset(fang_data, Group %in% c("ZMPBA", "ZMPIL", "ZMPJA")) 

# get rid of column 1 and 2
clean_teos <- select(teos_subset, abph1.20:zmm3.4)

trans_teos <- data.frame(t(clean_teos))

```


# Processing SNP data set
```{r}
# create a subset of the 3 necessary columns
snp_subset <- select(snp_data, SNP_ID, Chromosome, Position)

# combine the columns of the SNP data set and the maize data set
maize_combine <- cbind(snp_subset, trans_maize) 

# combine the columns of the SNP dataset and the teosinte data set
teos_combine <- cbind(snp_subset, trans_teos)

```

# Create 10 files for Maize data
```{r}
# ascending position order
maize_combine <- maize_combine %>% mutate(Position = as.numeric(Position))
ascend_maize <- arrange(maize_combine, Position)

# Filter by chromosome, ascending position order 

for (i in 1:10) {
  maize <- ascend_maize %>% filter(Chromosome == i) 
  file_name <- paste0("maize", i)  
  write.csv(maize, paste0(file_name, ".csv"), row.names = FALSE)
}


# descending position order
descend_maize <- arrange(maize_combine, desc(Position))
descend_maize[descend_maize == "?/?"] <- "-/-"

# Filter by chromosome, descending position order

for (i in 1:10) {
  dmaize <- descend_maize %>% filter(Chromosome == i) 
  file_name <- paste0("dmaize", i)  
  write.csv(dmaize, paste0(file_name, ".csv"), row.names = FALSE)
}

```



# Create 10 files for Teosinte data
```{r}
# ascending position order
teos_combine <- teos_combine %>% mutate(Position = as.numeric(Position))
ascend_teos <- arrange(teos_combine, Position)

# Filter by chromosome, ascending position order 
for (i in 1:10) {
  teosinte <- ascend_teos %>% filter(Chromosome == i) 
  file_name <- paste0("teos", i)  
  write.csv(teosinte, paste0(file_name, ".csv"), row.names = FALSE)
}

# descending position order
descend_teos <- arrange(teos_combine, desc(Position))
descend_teos[descend_teos == "?/?"] <- "-/-"

# Filter by chromosome, descending position order

for (i in 1:10) {
  dteosinte <- ascend_teos %>% filter(Chromosome == i) 
  file_name <- paste0("dteos", i)  
  write.csv(teosinte, paste0(file_name, ".csv"), row.names = FALSE)
}


```




# Visualization
```{r}
# Change chromosome number to factor for layout
maize_combine$Chromosome <- as.factor(maize_combine$Chromosome)

#SNPS per chromosome 
maize_snps_plot <- ggplot(maize_combine, aes(x= Chromosome)) +
  geom_bar(fill = "red") +
  labs(title = "Maize SNPs per Chromosome", x = "Chromosome", y = "SNPs")

maize_snps_plot
```
```{r}
# Change chromosome number to factor for layout

# Teosinte SNPS per chromosome 
teos_snps_plot <- ggplot(teos_combine, aes(x= Chromosome)) +
  geom_bar(fill = "lightblue") +
  labs(title = "Teosinte SNPs per Chromosome", x = "Chromosome", y = "SNPs")

teos_snps_plot


```
## SNP visualization
As the chromosome number increases, the number of SNPS per chromosome generally decreases. Chromosome 1 has the most about of SNPS for both the maize and teosinte. Chromosome 10 has the least number of SNPS out of all the chromosomes for both maize and teosinte. The SNP data on each chromosome is the same for maize and teosinte.

```{r}
# Missing data and Heterozygosity
## Maize data
homo_snps <- c("\\A/\\A", "\\C/\\C", "\\G/\\G", "\\T/\\T")
counts <- table(unlist(lapply(homo_snps, grepl, x = unlist(maize_combine))))
m_homozygotes <- counts[2]
vm_homozygotes <- as.vector(m_homozygotes)
total_snps <- 983*1576
m_heterozygotes <- total_snps - vm_homozygotes


## Teosinte data
homo_snps2 <- c("\\A/\\A", "\\C/\\C", "\\G/\\G", "\\T/\\T")
counts <- table(unlist(lapply(homo_snps2, grepl, x = unlist(teos_combine))))
t_homozygotes <- counts[2]
vt_homozygotes <- as.vector(t_homozygotes)
t_heterozygotes <- total_snps - vt_homozygotes

## Missing data
missing_snp<- length(grep("\\?/\\?", unlist(maize_combine)))

missing_snp2<- length(grep("\\?/\\?", unlist(teos_combine)))

maize_dataframe <- data.frame(Homozygous = vm_homozygotes, Heterozygous = m_heterozygotes, Missing = missing_snp)
maize_long <- gather(maize_dataframe, key = "variable", value = "value")

teos_dataframe <- data.frame(Homozygous = vt_homozygotes, Heterozygous = t_heterozygotes, Missing = missing_snp2)
teos_long <- gather(teos_dataframe, key = "variable", value = "value")


## Maize Plot
maize_snp_plot <- ggplot(maize_long, aes(x = variable, y = value)) +
    geom_bar(stat = "identity", fill = "red") + 
    labs(x = "Type", y = "Number of SNPS", fill = "Legend") +
    ggtitle("Maize SNP Information")

maize_snp_plot

## Teosinte Plot
teos_snp_plot <- ggplot(teos_long, aes(x = variable, y = value)) +
    geom_bar(stat = "identity", fill = "lightblue") +  
    labs(x = "Type", y = "Number of SNPS", fill = "Legend") +
    ggtitle("Teosinte SNP Information")

teos_snp_plot

```

# My Visualization
## Number of Maize samples vs Teosinte samples
```{r}
maize_samples <-  ncol(maize_combine) - 3

teos_samples <-  ncol(teos_combine) - 3

samples_df <- data.frame(Maize_Samples = maize_samples, Teosinte_Samples = teos_samples)
samples_long <- gather(samples_df, key = "Sample", value = "Count")

samples_plot <- ggplot(samples_long, aes(x = Sample, y = Count)) +
    geom_bar(stat = "identity", fill = "purple") +  
    labs(x = "Sample Type", y = "Number of Samples", fill = "Legend") +
    ggtitle("Maize Samples vs Teosinte Samples")

samples_plot
    
```

